Estimating Operator Intent\\ (Draft 4.0)

### 2015-09-30 intent.tex

\title{Estimating Operator Intent\\
(Draft 3.0)}

@@

Executive Summary to be written.

@@

This paper surveys state of the art technology to recognize
and predict exoskeleton
operator intent.
A companion paper surveys current operator/exoskeleton interfaces [survey.pdf].

@@

Predictive models of human motion, a major topic of robot learning from\\
observation/demonstration/imitation and computer animation,
could be the subject of a future white
paper~\cite{IEEE06913830}
\url{http://www.ri.cmu.edu/publication_view.html?pub_id=7891}

@@

Here is an example of a robot seeing and moving faster than a human, and
consistantly beating a human opponent in rock-paper-scissors.
\url{https://www.youtube.com/watch?v=Qb5UIPeFClM&feature=player_embedded}

@@

One customer of intent estimation is an exoskeleton control system.
Other possible customers are online operator situation awareness,
cognitive assistance
and physical guidance,
command and squad visualization, situation awareness,
coordination, and cognitive assistance,
after-action reports, and new operator training and
operator learning from practice and coaching,
We will focus on exoskeleton control in this paper.

@@

I have attached an Appendix with pointers to online behavior
recognition in related fields.

@@

It may be useful to estimate and predict operator intent in the form
of criteria the operator is ``optimizing''. These functions have
various names in different fields: utility or cost functions in economics
and decision sciences,
reward functions in reinforcement learning, and
optimization criteria and one-step-cost functions in optimal control.
Estimated utility functions may have a predetermined form and unknown
parameters and weights.
Predicting optimization criteria is known as ``inverse optimal control''
in optimal control and reinforcement learning.
\url{http://www.ri.cmu.edu/publication_view.html?pub_id=7800}

@@

Detection reaction time (the delay between an anticipated 
stimulus and the onset of movement) 
for the operators we are interested in (young adults)
is about
190ms for visual input,
160ms for auditory stimulus in a detection task,
It takes longer to recognize a signal from a set of ongoing events,
and even longer if the operator has to make a choice based on the signal
(go left or right, for example).
Reaction times for complex actions are longer that reaction times
for simple actions.
\url{https://en.wikipedia.org/wiki/Mental_chronometry}.

@@

We could have the control computer aware of the external world
available to the operator. These sensors are described in
a companion paper [technology.pdf].

@@

It is possible that external measurements of brain activity
(electroencephalography (EEG)) could detect stimuli processing, and
even more usefully, operator decision making.
It is also possible that implanted measurement devices could improve
greatly over external measurement devices in terms of specificity,
timing, and reliability of brain signal measurements.
\url{https://en.wikipedia.org/wiki/Electroencephalography}
Measurement of brain signals,
in the context of a wide range of other measurements,
might improve overall intent estimation.

@@

Muscle electrical activity can be measured by surface electrodes
(not very specific to particular muscles), needle electrodes (painful,
especially during movement), and implanted electrodes~\url{http://www.medscape.com/viewarticle/542356_2}.
Typical processing is to 
high-pass filter at 20 Hz, rectify, low-pass filter at 6 Hz,
offset by a small value, and amplify by some scaling factor
to get a muscle activation estimate~\cite{IEEE07139980}.
One can apply pattern recogntion to this signal~\cite{IEEE07222568}.

@@

@@

Operator-exoskeleton forces may be measured (forcemyography (FMG))
with various types of
force sensors such as single and multi-axis strain-gage load cells,
force sensitive resistors (FSRs), and pressure sensors built into an internal exoskeleton skin or at the
contacts (straps and cushions) of the exoskeleton attachment points.
Optical sensors and other distance sensors can measure deformation and
strain of the operator and between the operator and the exoskeleton.
Overall muscle ``pressure'' and other tissue parameters can be measured using
elastic straps around the arm, deformation pressure probes at various
points on the arm.
An estimate of individual muscle activity
may be available from tissue impedance measurements on the arm
or through the arm (including various forms of reflective and transmissive
ultrasound imaging tracking tissue parameters or implanted markers).
Some or all operator-exoskeleton forces may not be measured, and must
be estimated with some sort of Kalman filter.
Muscle activation can be measured (electromyography (EMG))
and force dynamics can be estimated to predict future operator-exoskeleton
forces.

@@

We recommend each exoskeleton controller
to be used by and optimized for a single operator.
A substantial investment in capturing the operators normal behavior,
operator training and learning, and controller customization should be made.
This paper discusses ways to model individual movement style
\url{http://humanmotion.ict.ac.cn/papers/2015P1_StyleTransfer/details.htm}
This paper discusses activity forecasting
\url{http://www.ri.cmu.edu/publication_view.html?pub_id=7250}

Predictive models of human motion, a major topic of robot learning from\\
observation/demonstration/imitation and computer animation,
could be the subject of a future white
paper~\cite{IEEE06913830}
\url{http://www.ri.cmu.edu/publication_view.html?pub_id=7891}

@@

We note that the operator will use more command channels that just
body movement, since
the operator controls more than the pose and motion of the
exoskeleton. What if there are ``pointing'' sensors or communication
devices that need to be aimed at or track an area of interest? What if
there are additional (physical or virtual) pan/tilt/zoom cameras
pointing to the side and rear or line of sight secure communications,
for example. What about other controls, such as power assist level or
thermal control? It is an open question
how an operator can naturally express intent and
control these additional degress of freedom?

\section{Discussion}

Discussion to be written.

@@

More Conclusions and Recommendations to be written.

@@

{\it 3D Gestural Interaction: The State of the Field,}
J. LaViola Jr.
\url{http://www.eecs.ucf.edu/~jjl/pubs/3dgesture.pdf}

@@

{\it Human Activity Prediction: Early Recognition of Ongoing Activities
from Streaming Videos,}
M. S. Ryoo
\url{http://michaelryoo.com/papers/iccv11_prediction_ryoo.pdf}

{\it Max-Margin Early Event Detectors,}
M. Hoai and F. De la Torre
\url{http://www.robots.ox.ac.uk/~minhhoai/papers/MMED_IJCV14.pdf}

@@

{\it Exploring the Trade-off Between Accuracy and Observational Latency in
Action Recognition,}
C. Ellis, S. Z. Masood, M. Tappen, J. LaViola Jr., R. Sukthankar
\url{http://www.cs.ucf.edu/~mtappen/pubs/2012/ellis-et-al-ijcv-2012.pdf}

{\it The Moving Pose: An Efficient 3D Kinematic Descriptor for Low-Latency
Action Recognition and Detection,}
M. Zanfir, M. Leordeanu, C. Sminchisescu
\url{http://www.cv-foundation.org/openaccess/content_iccv_2013/papers/Zanfir_The_Moving_Pose_2013_ICCV_paper.pdf}

{\it Real-time Classification of Dance Gestures from Skeleton Animation,}
M. Raptis, D. Kirovski, H. Hoppe
\url{http://vision.ucla.edu/papers/raptisKH11.pdf}

@@

{\it Activity Recognition on Streaming Sensor Data,}
N. Krishnan, D. Cook
\url{http://www.eecs.wsu.edu/~cook/pubs/pmc12.pdf}

{\it Macro-class Selection for Hierarchical kNN Classification of Inertial
Sensor Data,}
C. McCall, K. Reddy, M. Shah
\url{http://crcv.ucf.edu/papers/PECCS_2012.pdf}

{\it Audio-based Human Activity Recognition Using Non-Markovian Ensemble
Voting}
\url{http://www2.informatik.uni-freiburg.de/~spinello/storkROMAN12.pdf}

>>>>

\title{Estimating Operator Intent\\
(Draft 4.0)}

@@

\begin{enumerate}
\item
We recommend using a Bayesian {\bf (probabilistic) framework}
for intent recognition and prediction.
\item
We recommend developing an intent prediction system that can use
as {\bf many sensors and measurements} of the operator as are possible.
\item
We recommend each exoskeleton controller
to be used by and {\bf optimized for a single operator.}
A substantial investment in capturing the operators normal behavior,
operator training and learning, and controller customization should be made.
\end{enumerate}

@@

This paper surveys state of the art technology to recognize
and predict exoskeleton
operator intent.
A companion paper surveys current operator/exoskeleton interfaces.\\
\url{http://www.cs.cmu.edu/~cga/exo/survey.pdf}

@@

Predictive models of human motion, a major topic of robot learning from\\
observation/demonstration/imitation and computer animation,
could be the subject of a future white
paper.\\
%~\cite{IEEE06913830}
\url{http://ieeexplore.ieee.org/xpl/login.jsp?tp=&arnumber=6913830}\\
\url{http://www.ri.cmu.edu/publication_view.html?pub_id=7891}

@@

Here is an example of a robot seeing and moving faster than a human, and
consistantly beating a human opponent in rock-paper-scissors.\\
\url{https://www.youtube.com/watch?v=Qb5UIPeFClM&feature=player_embedded}

@@

One customer of intent estimation is an exoskeleton control system.
Other possible customers are online operator situation awareness,
cognitive assistance
and physical guidance;
command and squad visualization, situation awareness,
coordination, and cognitive assistance;
after-action reports; and new operator training and
operator learning from practice and coaching,
We will focus on exoskeleton control in this paper.

@@

We have attached an Appendix with pointers to online behavior
recognition in related fields.

@@

It may be useful to estimate and predict operator intent in the form
of criteria the operator is ``optimizing''. These functions have
various names in different fields: utility or cost functions in economics
and decision sciences,
reward functions in reinforcement learning, and
optimization criteria and one-step-cost functions in optimal control.
Estimated utility functions may have a predetermined form and unknown
parameters and weights.
Predicting optimization criteria is known as ``inverse optimal control''
in optimal control and reinforcement learning.\\
\url{http://www.ri.cmu.edu/publication_view.html?pub_id=7800}

@@

Detection reaction time (the delay between an anticipated 
stimulus and the onset of movement) 
for the operators we are interested in (young adults)
is about
190ms for visual input and
160ms for auditory stimulus in a detection task.
It takes longer to recognize a signal from a set of ongoing events,
and even longer if the operator has to make a choice based on the signal
(go left or right, for example).
Reaction times for complex actions are longer that reaction times
for simple actions.\\
\url{https://en.wikipedia.org/wiki/Mental_chronometry}.

@@

We could have the control computer aware of the external world
available to the operator. These sensors are described in
a companion paper.\\
\url{http://www.cs.cmu.edu/~cga/exo/technology.pdf}

@@

It is possible that external measurements of brain activity
(electroencephalography (EEG)) could detect stimuli processing, and
even more usefully, operator decision making.
It is also possible that implanted measurement devices could improve
greatly over external measurement devices in terms of specificity,
timing, and reliability of brain signal measurements.\\
\url{https://en.wikipedia.org/wiki/Electroencephalography}\\

You can take two views of using brain signals to control robots or
exoskeletons.
The first view is that there is great promise, based on demos like
the one at the recent World Cup:\\
\url{https://www.youtube.com/watch?t=3&v=myNRcDIgfjw}\\
\url{www.livescience.com/51940-mind-controlled-exoskeleton-robot.html}\\
\url{http://www.ibtimes.co.uk/scientists-develop-mind-controlled-robotic-exoskeleton-that-uses-leds-help-paraplegics-walk-1516834}\\

The second view is that the field is full of crap. In the above demo,
the user looked at LEDs in order to get a measurable brain signal.
This means that one could have just tracked where the user was looking
(no brain signals used at all) to figure out what to do.
In general the bit rate coming out of any brain-machine interface is
very very low, so decoding the signal is going to be very very slow
(probably longer than the reaction time of an athletic subject).

Nevertheless, a middle of the road view might be beneficial. It is possible
that brain signals,
in the context of a wide range of other measurements,
might improve overall intent estimation.
It may be that brain signals are useful to tell us {\bf when} a behavior
is happening, if they are not very useful for telling us {\bf what}
behavior is happening.

@@

Muscle electrical activity can be measured by surface electrodes
(not very specific to particular muscles), needle electrodes (painful,
especially during movement), and implanted electrodes.\\
\url{http://www.medscape.com/viewarticle/542356_2}.\\
Typical processing is to 
high-pass filter at 20 Hz, rectify, low-pass filter at 6 Hz,
offset by a small value, and amplify by some scaling factor
to get a muscle activation estimate.\\
\url{http://ieeexplore.ieee.org/xpl/login.jsp?tp=&arnumber=7139980}\\
One can apply pattern recogntion to this signal.\\
\url{http://ieeexplore.ieee.org/xpl/login.jsp?tp=&arnumber=7222568}\\

@@

Off the shelf electronics:\\
\url{http://www.advancertechnologies.com/p/muscle-sensor-v3.html}

@@

Operator-exoskeleton forces may be measured (forcemyography (FMG))
with various types of
force sensors such as single and multi-axis strain-gage load cells,
force sensitive resistors (FSRs), and pressure sensors built into an internal exoskeleton skin or at the
contacts (straps and cushions) of the exoskeleton attachment points.
Optical sensors and other distance sensors can measure deformation and
strain of the operator and between the operator and the exoskeleton.
Overall muscle ``pressure'' and other tissue parameters can be measured using
elastic straps around the arm, deformation pressure probes at various
points on the arm.
An estimate of individual muscle activity
may be available from tissue impedance measurements on the arm
or through the arm (including various forms of reflective and transmissive
ultrasound imaging tracking tissue parameters or implanted markers).
Some or all operator-exoskeleton forces may not be measured, and must
be estimated with some sort of Kalman filter.
EMG and FMG can be usefully combined.

\url{http://ieeexplore.ieee.org/xpl/login.jsp?tp=&arnumber=6913842}\\

\begin{verbatim}
[8] K. Kong and M. Tomizuka, “A Gait Monitoring System Based on Air
Pressure Sensors Embedded in a Shoe,” IEEE/ASME Transactions on
Mechatronics, vol. 14, pp. 358–370, 2009.
\end{verbatim}

@@

We recommend each exoskeleton controller
to be used by and optimized for a single operator.
A substantial investment in capturing the operators normal behavior,
operator training and learning, and controller customization should be made.
This paper discusses ways to model individual movement style.\\
\url{http://humanmotion.ict.ac.cn/papers/2015P1_StyleTransfer/details.htm}\\
This paper discusses activity forecasting.\\
\url{http://www.ri.cmu.edu/publication_view.html?pub_id=7250}

Predictive models of human motion, a major topic of robot learning from\\
observation/demonstration/imitation and computer animation,
could be the subject of a future white
paper.\\
\url{http://ieeexplore.ieee.org/xpl/login.jsp?tp=&arnumber=6913830}\\
\url{http://www.ri.cmu.edu/publication_view.html?pub_id=7891}

@@

We note that the operator will use more command channels that just
body movement, since
the operator controls more than the pose and motion of the
exoskeleton. What if there are ``pointing'' sensors or communication
devices that need to be aimed at or track an area of interest? What if
there are additional (physical or virtual) pan/tilt/zoom cameras
pointing to the side and rear or line of sight secure communications,
for example. What about other controls, such as power assist level or
thermal control? It is an open question
how an operator can naturally express intent and
control these additional degress of freedom.

@@

@@

{\it 3D Gestural Interaction: The State of the Field,}\\
J. LaViola Jr.\\
\url{http://www.eecs.ucf.edu/~jjl/pubs/3dgesture.pdf}

@@

{\it Human Activity Prediction: Early Recognition of Ongoing Activities
from Streaming Videos,}\\
M. S. Ryoo\\
\url{http://michaelryoo.com/papers/iccv11_prediction_ryoo.pdf}

{\it Max-Margin Early Event Detectors,}\\
M. Hoai and F. De la Torre\\
\url{http://www.robots.ox.ac.uk/~minhhoai/papers/MMED_IJCV14.pdf}

@@

{\it Exploring the Trade-off Between Accuracy and Observational Latency in
Action Recognition,}\\
C. Ellis, S. Z. Masood, M. Tappen, J. LaViola Jr., R. Sukthankar\\
\url{http://www.cs.ucf.edu/~mtappen/pubs/2012/ellis-et-al-ijcv-2012.pdf}

{\it The Moving Pose: An Efficient 3D Kinematic Descriptor for Low-Latency
Action Recognition and Detection,}\\
M. Zanfir, M. Leordeanu, C. Sminchisescu\\
\url{http://www.cv-foundation.org/openaccess/content_iccv_2013/papers/Zanfir_The_Moving_Pose_2013_ICCV_paper.pdf}

{\it Real-time Classification of Dance Gestures from Skeleton Animation,}\\
M. Raptis, D. Kirovski, H. Hoppe\\
\url{http://vision.ucla.edu/papers/raptisKH11.pdf}

@@

{\it Activity Recognition on Streaming Sensor Data,}\\
N. Krishnan, D. Cook\\
\url{http://www.eecs.wsu.edu/~cook/pubs/pmc12.pdf}

{\it Macro-class Selection for Hierarchical kNN Classification of Inertial
Sensor Data,}\\
C. McCall, K. Reddy, M. Shah\\
\url{http://crcv.ucf.edu/papers/PECCS_2012.pdf}

{\it Audio-based Human Activity Recognition Using Non-Markovian Ensemble
Voting}\\
\url{http://www2.informatik.uni-freiburg.de/~spinello/storkROMAN12.pdf}
